---
title: "SW CellMixture MAESTER"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "CairoPNG")
```

## Reproduction of cell mixture results using Seq-Well data

This is a reproduction using the MAESTER script 3.1_SW_CellLineMix_variants.R.

Available here: https://github.com/petervangalen/MAESTER-2021/blob/main/3_Cell_line_mixes_variants/3.1_SW_CellLineMix_variants.R

## Loading necessary packages.

We load packages necessary for the MAESTER script.

```{r Packages}
print("Libraries for MAESTER.")
options(stringsAsFactors = FALSE)
options(scipen = 999)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(Matrix))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(ggrastr))
```

## Loading the data using the MAESTER script.

We load the data using the MAESTER script. This requires the use of multiple functions and objects.

```{r warning = FALSE, dev = "CairoPNG"}
popcol.tib <- read_excel("MAESTER_colors.xlsx")
mycol.ch <- popcol.tib$hex
names(mycol.ch) <- popcol.tib$name
# Custom Functions
cutf <- function(x, f=1, d="/") sapply(strsplit(x, d), function(i) paste(i[f], collapse=d))
computeAFMutMatrix <- function(SE){
  cov <- assays(SE)[["coverage"]]+ 0.000001
  ref_allele <- as.character(rowRanges(SE)$refAllele)
  getMutMatrix <- function(letter){
    mat <- (assays(SE)[[paste0(letter, "_counts_fw")]] + assays(SE)[[paste0(letter, "_counts_rev")]]) / cov
    rownames(mat) <- paste0(as.character(1:dim(mat)[1]), "_", toupper(ref_allele), ">", letter)
    return(mat[toupper(ref_allele) != letter,])
  }
  rbind(getMutMatrix("A"), getMutMatrix("C"), getMutMatrix("G"), getMutMatrix("T"))
}

# Loading the scRNAseq data.
seu <- readRDS("SW_CellLineMix_Seurat_Keep.rds")

# Loading the MAEGATK results.
maegatk_all.rse <- readRDS("SW_CellLineMix_All_mr3_maegatk.rds")

# Intersecting the cells. We retain only cells that are in both data sets.
common.cells <- intersect(colnames(seu), colnames(maegatk_all.rse))
seu <- seu[,common.cells]
maegatk.rse <- maegatk_all.rse[,common.cells]

# Calculating the allele frequency per cell and possible variant.
af.dm <- data.matrix(computeAFMutMatrix(maegatk.rse))*100

# Organizing the cellular meta data in a tibble.
cells.tib <- tibble(cell = common.cells,
                    orig.ident = seu$orig.ident,
                    CellType_RNA = seu$CellType,
                    Mean_Cov = maegatk.rse$depth)

# Separating the cells into distinct subgroups.
CellSubsets.ls <- list(unionCells = cells.tib$cell,
                       K562 = filter(cells.tib, CellType_RNA == "K562")$cell,
                       BT142 = filter(cells.tib, CellType_RNA == "BT142")$cell)

# Calculating the average allele frequency per variant.
mean_af.ls <- lapply(CellSubsets.ls, function(x) rowMeans(af.dm[,x]))
names(mean_af.ls) <- paste0("mean.af.", names(mean_af.ls))

# Calculating the average coverage per variant.
mean_cov.ls <- lapply(CellSubsets.ls, function(x) rowMeans(assays(maegatk.rse)[["coverage"]][,x])[as.numeric(cutf(rownames(af.dm), d = "_"))])
names(mean_cov.ls) <- paste0("mean.cov.", names(mean_cov.ls))

# Calculating the allele frequency quantile per variant.
quantiles <- c("q01" = 0.01, "q10" = 0.1, "q50" = 0.5, "q90" = 0.9, "q99" = 0.99)
quantiles.ls <- lapply(quantiles, function(x) lapply(CellSubsets.ls, function(y) apply(af.dm[,y], 1, quantile, x) ))
assays.ls <- lapply(maegatk.rse@assays$data, function(x) as.matrix(x))
qual.num <- sapply(rownames(af.dm), function(x){
  pos <- as.numeric(cutf(x, d = "_"))
  mut <- cutf(x, d = ">", f = 2)
  covered_fw <- assays.ls[[str_c(mut, "_counts_fw")]][pos,] > 0
  qual_fw <- assays.ls[[str_c(mut, "_qual_fw")]][pos, covered_fw]
  covered_rev <- assays.ls[[str_c(mut, "_counts_rev")]][pos,] > 0
  qual_rev <- assays.ls[[str_c(mut, "_qual_rev")]][pos, covered_rev]
  qual <- mean(c(qual_fw, qual_rev))
  return(qual)
})

# Organizing the results in a variant-wise tibble.
vars.tib <- as_tibble(do.call(cbind, c(mean_af.ls, mean_cov.ls, unlist(quantiles.ls, recursive = F))), rownames = "var")
vars.tib <- add_column(vars.tib, quality = qual.num, .before = 2)
```

## Selecting the Variants of Interest

Now, we select the variants of interest.

```{r dev = "CairoPNG"}
# Identifying variants of interest by subsetting the variant-wise tibble.
voi.ch <- vars.tib %>% filter(mean.cov.unionCells > 20,
                              quality > 30,
                              q10.unionCells < 10,
                              q90.unionCells > 90) %>% .$var
print(voi.ch)
```

## Plotting the mtVOIs as a heatmap

```{r dev = "CairoPNG"}
# Generate matrices with coverage, allele frequency and reference / variant reads
cov_voi.mat <- assays(maegatk.rse)[["coverage"]][as.numeric(cutf(voi.ch, d = "_")),]
af_voi.mat <- af.dm[voi.ch,]

# Add coverage and allele frequency info from variants of interest to cells.tib
for (voi in voi.ch) {
  cells.tib <- cells.tib %>%
    left_join(as_tibble(assays(maegatk.rse)[["coverage"]][as.numeric(cutf(voi, d = "_")),], rownames = "cell"), by = "cell") %>%
    left_join(as_tibble(af.dm[voi,], rownames = "cell"), by = "cell") %>%
    rename(value.x = str_c("cov_", str_replace(voi, ">", ".")), value.y = str_c("af_", str_replace(voi, ">", ".")))
}

# Calculating the support of each cell.
supporting_calls <- function(voi, maegatk) {
    # From the variant, determine position, reference and mutant allele
    pos <- as.numeric(cutf(voi, d = "_"))
    ref <- cutf(voi, d = "_|>|\\.", f = 2)
    mut <- cutf(voi, d = ">|\\.", f = 2)
    # Put all the counts for this variant from the maegatk object in a data frame
    counts.df <- do.call(cbind, lapply(assays(maegatk)[grepl("counts", names(assays(maegatk)))], function(x) x[pos,]))
    # Add up reads supporting reference or mutated allele
    support.df <- data.frame(ref = rowSums(counts.df[,grepl(paste0("^", ref), colnames(counts.df))]),
                             mut = rowSums(counts.df[,grepl(paste0("^", mut), colnames(counts.df))]))
    support.df
}
supporting_calls.ls <- lapply(voi.ch, function(x) supporting_calls(x, maegatk.rse))
names(supporting_calls.ls) <- voi.ch
combined_supporting_calls.df <- do.call(cbind, supporting_calls.ls)

supporting_calls.tib <- tibble(cell = rownames(combined_supporting_calls.df),
    K562_supporting_calls = rowSums(combined_supporting_calls.df[,c("709_G>A.mut", "1888_G>A.mut", "1420_T>C.mut", "2141_T>C.mut", "9117_T>C.mut", "7990_C>T.ref")]),
    BT142_supporting_calls = rowSums(combined_supporting_calls.df[,c("709_G>A.ref", "1888_G>A.ref", "1420_T>C.ref", "2141_T>C.ref", "9117_T>C.ref", "7990_C>T.mut")]))
# Combine information
cells.tib <- left_join(cells.tib, supporting_calls.tib, by = "cell")

cells.tib <- cells.tib %>% mutate(CellType_MT = ifelse(K562_supporting_calls + BT142_supporting_calls < 30, yes = "NoCoverage", no =
    ifelse(K562_supporting_calls > 3 & K562_supporting_calls > 10*BT142_supporting_calls, yes = "K562", no =
    ifelse(BT142_supporting_calls > 3 & BT142_supporting_calls > 10*K562_supporting_calls, yes = "BT142", no = "Contaminated")))) %>%
    mutate(CellType_MT = factor(CellType_MT, levels = c("Contaminated", "BT142", "K562", "NoCoverage")))

# Selecting cells that have sufficient coverage for all the variants of interest.
cells.ch <- cells.tib %>% filter(CellType_MT %in% c("K562", "BT142")) %>%
    select(cell, starts_with("cov_")) %>%
    filter(apply(.[,-1], 1, function(x) all(x > 3))) %>%
    .$cell

# Generating the annotation object for the heatmap.
ha <- HeatmapAnnotation(CellType_RNA = cells.tib$CellType_RNA[cells.ch],
                        col = list(CellType_RNA = c("K562" = "#7BF581", "BT142" = "#8E87F5")))

# Visualize: clustered VAFs
Heatmap(af_voi.mat[,cells.ch], name = "VAF", col=c("#FCFCFC","#FFEDB0","#FFDF5F","#FEC510","#FA8E24","#F14C2B","#DA2828","#BE2222","#A31D1D"), show_row_names = TRUE, show_column_names = FALSE, cluster_columns = TRUE, cluster_rows = FALSE, top_annotation = ha, border = TRUE,
               heatmap_legend_param = list(border = "black", at = c(0,25,50,75,100)), use_raster = TRUE)
```

