---
title: "BPDCN BoneMarrow SIGURD"
author: |
  Martin Grasshoff<sup>1</sup>, Ivan G. Costa<sup>1</sup>

affiliations: |
  1. Institute for Computational Genomics, Faculty of Medicine, RWTH Aachen University, Aachen, 52074 Germany

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BPDCN BoneMarrow SIGURD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \usepackage[utf8]{inputenc}  # Optional, if needed for LaTeX output.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "CairoPNG")
```

## Analysis of Blastic Plasmacytoid Dendritic Cell Neoplasm (BPDCN)

Here, we compare the analysis of cells from a patient with BPDCN.

The original script is available here: https://github.com/petervangalen/MAESTER-2021/blob/main/4_CH_sample/4.2_Variant_Selection.R

In this setup, the cells are all from a single donor and do not have to be separated into categories. 

Then small clones that might characterize the clonal lineages in the sample are selected.

## Loading necessary packages.

We load the data from MAESTER. 

```{r Packages, warning = FALSE}
print("Libraries for SIGURD.")
suppressPackageStartupMessages(library(sigurd))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
```

## Loading the data using SIGURD.

```{r Loading_SIGURD, warning = FALSE}
# The design matrix contains information for the genotyping data.
genotyping <- LoadingMAEGATK_typewise(patient = "BPDCN712", samples_file = "data/MAESTER_Reproduction.csv", type_use = "Amplicon_MT", verbose = FALSE)

# Loading the scRNA-seq data.
scrna <- readRDS("data/BPDCN712_Seurat_with_TCR_Renamed.rds")
```

## Generating a block list

Variants that are also detected in the cell mixture data is treated as possible false positives. They are used as a blacklist and are removed from the results.

```{r warning = FALSE}
# Loading the TenX Cell Mixture Genotyping.

genotyping_tenx <- load_object("data/TenX_CellMixture_Genotyping.rds.lz4")
blocklist <- AllelFrequencyFoldChange(genotyping_tenx, group_of_interest = "CellType", group1 = "K562", group2 = "BT142", maximum_foldchange = 5, minimum_coverage = 5, minimum_allele_freq = 0.001, maximum_allele_freq = 0.999)$Variant

# We add the variant chrM_1583_A_G by hand. It is identified as misleading based on downstream analysis.
blocklist <- c(blocklist, "chrM_1583_A_G")
```
## Selecting the Variants of Interest

Now, we select the variants of interest from the loaded data.

```{r warning = FALSE}
voi.ch.sigurd <- VariantSelection_TopCells(genotyping, min_coverage = 5, quantiles = 0.9, thresholds = 0, top_cells = 10, top_VAF = 0.5, min_quality = 30, remove_nocall = FALSE, verbose = FALSE)
voi.ch.sigurd <- voi.ch.sigurd[!voi.ch.sigurd %in% blocklist]
print(voi.ch.sigurd)
```

## Visualisation using SIGURD

```{r Vis_SIGURD, fig.width = 6, fig.height = 4, warning = FALSE}
genotyping <- Filtering(SE = genotyping, cells_include = colnames(scrna))
colData(genotyping)$CellType <- scrna$CellType
HeatmapVoi(SE = genotyping, voi = voi.ch.sigurd, annotation_trait = "CellType", sort_cells = TRUE, remove_empty_cells = TRUE, minimum_allele_freq = 0.01)
```
