---
title: "MPN SIGURD"
author: "Martin Grasshoff<sup>1</sup>, Ivan G. Costa<sup>1</sup>"
affiliations: "1. Institute for Computational Genomics, Faculty of Medicine, RWTH Aachen University, Aachen, 52074 Germany"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MPN SIGURD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - '\usepackage[utf8]{inputenc}'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading necessary packages.

```{r Packages, warning = FALSE}
suppressPackageStartupMessages(library(sigurd))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(EnhancedVolcano))
```
## Analysis of MPN data

Here, we illustrate how to analysis genotyping data from MPN samples. The JAK2V617F mutation was genotyped for 6 MPN samples and 3 healthy controls.

This setup makes this analysis different from the previous analyses. Previously, only one sample was analysed at a time. Now, we load and analyse the samples simultaneously.

## Loading MPN data
Now, we load the JAK2V617F data and plot the mutated cells on the UMAP.

Since we are loading multiple samples, this code is more complicated than the code for the MAESTER SW sample.

```{r MPN, warning = FALSE}
# Loading the scRNA-seq data.
scrna <- load_object("/data/MPN/exp/scRNA/MPN_mutations/SIGURD_paper/sigurd/data/scrna_MPN.rds.lz4")

# Loading the genotyping data.
genotyping  <- LoadingVarTrix_typewise(patient = "MPN", samples_file = "/data/MPN/exp/scRNA/MPN_mutations/SIGURD_paper/sigurd/data/DesignMatrix_VarTrix.csv", min_reads = 0, vcf_path = "/data/MPN/exp/scRNA/MPN_mutations/SIGURD_paper/sigurd/data/CosmicSubset_JAK2V617F.vcf", type_use = "scRNAseq_Somatic", min_cells = 0, verbose = FALSE)
amplicon <- LoadingVarTrix_typewise(patient = "MPN", samples_file = "/data/MPN/exp/scRNA/MPN_mutations/SIGURD_paper/sigurd/data/DesignMatrix_Amplicon.csv", min_reads = 0, vcf_path = "/data/MPN/exp/scRNA/MPN_mutations/SIGURD_paper/sigurd/data/CosmicSubset_UCSC_JAK2V617F.vcf", type_use = "scRNAseq_Amplicon", min_cells = 0, verbose = FALSE)
genotyping  <- AmpliconSupplementing(scRNAseq = genotyping, amplicon = amplicon, verbose = FALSE)

# Filtering the genotyping object to remove false positives.
genotyping <- Filtering(genotyping, min_cells_per_variant = 0, fraction_threshold = 0.21, cells_include = colnames(scrna), min_variants_per_cell = 1, reject_value = "NoCall", verbose = FALSE)
```

## Combining Seurat and SIGURD

Now, we add the genotyping information to the Seurat object.

```{r warning = FALSE}
scrna <- SetVariantInfo(SE = genotyping, seurat_object = scrna)
```

## Visualisation

We visualise the genotyping information on the UMAP.

```{r warning = FALSE, fig.width = 4, fig.height = 4}
# Plotting the mutated cells on the UMAP.
scrna$JAK2_p.V617F_c.1849G.T_consensus <- factor(scrna$JAK2_p.V617F_c.1849G.T_consensus, levels = c("Ref", "Alt"))
jak2_cells <- sum(scrna$JAK2_p.V617F_c.1849G.T_consensus == "Alt", na.rm = TRUE)
ref_cells <- sum(scrna$JAK2_p.V617F_c.1849G.T_consensus == "Ref", na.rm = TRUE)
all_cells <- ncol(scrna)
p <- DimPlot(scrna, group.by = "JAK2_p.V617F_c.1849G.T_consensus", reduction = "INTE_UMAP", order = TRUE, pt.size = 1, raster = FALSE, na.value = "grey95", cols = c(Alt = "#aa0051", Ref = "#79b938")) +
  ggtitle(paste0("Alt: ", jak2_cells, ", Ref: ", ref_cells, ", All: ", all_cells)) +
  xlab("UMAP 1") + ylab("UMAP 2") + scale_color_manual(breaks = c("Alt", "Ref"), values = c("#aa0051", "#79b938"), na.value = "grey95") +
  theme(legend.position = "top")
print(p)
```

## We determine DEGs between the JAK2V617F and WT cells.

Now, we determine the DEGs between the JAK2V617F cells and the wild type cells for the Orthochromatic Erythroblasts.

```{r DEGs_OrthoE_JAK2V617F, fig.width = 6, fig.height = 6, warning = FALSE}
scrna_orthoe <- subset(scrna, CellType == "Orthochromatic Erythroblasts" & Condition != "HC")
degs <- FindMarkers(scrna_orthoe, group.by = "JAK2_p.V617F_c.1849G.T_consensus", ident.1 = "Alt", ident.2 = "Ref", logfc.threshold = 0)
degs <- data.frame(degs, gene = rownames(degs))
degs_up   <- nrow(subset(degs, avg_log2FC >   0.25 & p_val_adj < 0.05))
degs_down <- nrow(subset(degs, avg_log2FC <  -0.25 & p_val_adj < 0.05))
EnhancedVolcano(degs, x = "avg_log2FC", y = "p_val_adj", lab = degs$gene, FCcutoff = 0.25, pCutoff = 0.05, title = "OrthoE JAK2V617F Pos. vs. Neg.",subtitle = NULL)
```

## Combining the JAK2V617F data with the MT data.

We combine the JAK2V617F SIGURD object with the MT object. Cells and variants in both objects are combined.

To save time, we are only processing a single sample. All other samples are processed equally. Since the genotyping object is a list, it is easily processed using parallel computing.

First, we load and filter the mitochondrial variant information.

```{r Combining, warning = FALSE}
# Loading the data. 
genotyping_mt <- LoadingMAEGATK_typewise(patient = "MPN1", samples_file = "/data/MPN/exp/scRNA/MPN_mutations/SIGURD_paper/sigurd/data/DesignMatrix_MT.csv", min_cells = 0, cells_include = colnames(scrna), type_use = "scRNAseq_MT", verbose = FALSE)
genotyping_mt <- Filtering(genotyping_mt, min_variants_per_cell = 1, min_cells_per_variant = 2, fraction_threshold = 0.05, cells_include = colnames(scrna), verbose = FALSE)
```

## MPN VOIs and Diversity

Now, we select the mtVOIs. We then determine the clonal lineages and plot the results on a heatmap.

We subset the cells to only include the erythroid cells.

```{r MPN_VOI, fig.width = 6, warning = FALSE, dev = "CairoPNG"}
# Subset to erythroid cells.
erythroid_cells <- subset(scrna, annotation == "ErythroidCells")
genotyping_mt <- Filtering(genotyping_mt, cells_include = colnames(erythroid_cells))

# Selecting the mtVOIS.
mt_vois <- VariantSelection_Quantile(genotyping_mt, min_coverage = 1, verbose = FALSE)

# Determining the clonal lineages.
genotyping_mt <- ClonalDefinition(se = genotyping_mt, variants_ls = list(mt_vois), verbose = FALSE)
```

## Visualisation of mtVOIs and Clonal Lineages

We now plot the variants of interest and the clonal lineages.

```{r warning = FALSE, fig.width = 6, fig.height = 4, dev = "CairoPNG"}
# Generating a heatmap.
colors_lineage <- c("C1" = "#563e00", "C2" = "#dec234", "C3" = "#e23856", "C4" = "#7f0073", "C5" = "#006639", "C6" = "#b1db8e", "C7" = "#00348f", "Negative" = "black")
ha <- columnAnnotation(Clones = colData(genotyping_mt)[,"Clones"],
                       col = list(Clones = colors_lineage),
                       annotation_name_gp = gpar(fontsize = 10, fontface = "plain"),
                       annotation_legend_param = list(Clones = list(nrow = 1, title = "")))
fraction <- as.matrix(assays(genotyping_mt)[["fraction"]])[mt_vois, ]
hm <- Heatmap(fraction,
  name = "VAF",
  row_title_gp = gpar(fontsize = 12),
  row_names_gp = gpar(fontsize = 10),
  row_names_max_width = unit(16, "cm"),
  column_title = "MPN1",
  column_title_gp = gpar(fontsize = 12),
  column_names_gp = gpar(fontsize = 10),
  col = colorRamp2(seq(0, round(max(fraction, na.rm = TRUE)), length.out = 9), c("#FCFCFC", "#FFEDB0", "#FFDF5F", "#FEC510", "#FA8E24", "#F14C2B", "#DA2828", "#BE2222", "#A31D1D")),
  show_column_dend = FALSE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  cluster_columns = TRUE,
  cluster_rows = FALSE,
  heatmap_legend_param = list(border = "#000000", grid_height = unit(10, "mm"),
  direction = "horizontal"),
  bottom_annotation = ha,
  border = TRUE,
  use_raster = TRUE,
  show_heatmap_legend = TRUE)
draw(hm, annotation_legend_side = "bottom")
```

## Calculating clonal diversity.

We calculate the clonal diversity for the sample MPN1 as an example.

```{r Diversity, warning = FALSE, dev = "CairoPNG"}
clonal_diversity <- ClonalDiversity(genotyping_mt, grouping = "Clones", diversity_measure = "EffectiveSpecies", verbose = FALSE)
print(paste0("Clonal Diversity in Effective Number of Species for MPN1: ", clonal_diversity))
```
