---
title: "SiGURD vignette"
author: 
- name: Martin Grasshoff
  affilitation:
     - Institute for Computational Genomics,
       Faculty of Medicine,
       RWTH Aachen University,
       Aachen, 52074 Germany
- name: Ivan G. Costa
  affiliation:
     - Institute for Computational Genomics,
       Faculty of Medicine,
       RWTH Aachen University,
       Aachen, 52074 Germany
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r vignetteLibraries, warning = FALSE, message = FALSE, output = FALSE}
suppressPackageStartupMessages(require(sigurd))
suppressPackageStartupMessages(require(SummarizedExperiment))
suppressPackageStartupMessages(require(VariantAnnotation))
```

**R version**: `r R.version.string`

**Package version**: `r packageVersion("CrossTalkeR")`


# Your input file.

```{r InputFile, warning = FALSE, message = FALSE, output = FALSE}
sample_path <- system.file("extdata", "Input_Example_local.csv", package = "sigurd")
sample_file <- read.csv(sample_path)
print(sample_file)

```
# Your vcf files.

These files are needed for VarTrix and not MGATK/MAEGATK. Since MAEGATK only analyses the mitochondrial genome, you only need to decide on a chromosomal prefix for your data. See the loading of data below.

```{r InputVCF, warning = FALSE, message = FALSE, output = FALSE}
vcf_path <- system.file("extdata", "ALFA_subset_MAF2_prefix.vcf", package = "sigurd")
vcf <- readVcf(vcf_path)
vcf_info <- info(vcf)
print(vcf)
print(vcf_info)

vcf_path_mt <- system.file("extdata", "chrM_Input_VCF_NoMAF_Filtering.vcf", package = "sigurd")
vcf_mt <- readVcf(vcf_path_mt)
vcf_mt_info <- info(vcf_mt)
print(vcf_mt)
print(vcf_mt_info)

```

# Loading and filtering the input data.

We load the data per patient and merge all the associated samples automatically.
In the input file, you have to include which software tool was used for the analysis. 
The source can either be vartrix or maegatk/mgatk. The respective loading function will then only load the files intended for it.
The types of data available are:
- scRNAseq_Somatic: the standard 10X results analysed for somatic variants.
- scRNAseq_MT: the standard 10X results analysed for MT variants. 
- Amplicon_Somatic: amplicon data analysed for somatic variants.
- Amplicon_MT: amplicon data analysed for MT variants.

Since the MT results are denser, they take longer to load. 

```{r LoadData, warning = FALSE, message = FALSE, output = FALSE}

Sample1_scRNAseq_Somatic <- LoadingVarTrix_typewise(samples_file = sample_path, vcf_path = vcf_path, patient = "Sample1", type_use = "scRNAseq_Somatic")

Sample1_scRNAseq_MT <- LoadingVarTrix_typewise(samples_file = sample_path, vcf_path = vcf_path_mt, patient = "Sample1", type_use = "scRNAseq_MT")

Sample1_combined <- CombineSEobjects(se_somatic = Sample1_scRNAseq_Somatic, se_MT = Sample1_scRNAseq_MT, suffixes = c("_somatic", "_MT"))
rm(Sample1_scRNAseq_Somatic, Sample1_scRNAseq_MT)
Sample1_combined <- Filtering(Sample1_combined, min_cells_per_variant = 2, fraction_threshold = 0.05)
Sample1_combined <- VariantBurden(Sample1_combined)

```

# Determing MT variants of interest.

This thresholding was adapted from Miller et al.
https://github.com/petervangalen/MAESTER-2021
https://www.nature.com/articles/s41587-022-01210-8

The heatmap needs some time to plot, since the cells are clustered. 

```{r VariantsOfInterest, warning = FALSE, message = FALSE, output = FALSE}
voi_ch <- VariantQuantileThresholding(SE = Sample1_combined, min_coverage = 2, quantiles = c(0.1, 0.9), thresholds = c(0.1, 0.9))
hm <- HeatmapVoi(SE = Sample1_combined, voi = voi_ch)
print(hm)
```

# Association of Variants

Using Fisher's Exact test, we find co-present variants. You can also use the correlation between variants.
For this, we combine the somatic and the MT results.
Since the possible number of tests/correlations is quite large, you can use multiple cores to perform the calculations.


```{r VariantAssociation, warning = FALSE, message = FALSE, output = FALSE}
Sample1_split_rows <- RowWiseSplit(Sample1_combined, remove_nocalls = FALSE)

results_fishertest  <- VariantWiseFisherTest(Sample1_split_rows, n_cores = 8)
rm(Sample1_split_rows)

variant_association_heatmap <- VariantFisherTestHeatmap(results_fishertest, patient = "Sample1", min_alt_cells = 3)
print(variant_association_heatmap)
```
